---
title: "Exercise sheet: Day 2"
author: "Vangelis Theodorakis, Fatemeh Behjati, Julien Gagneur, Marcel Schulz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{CeDoSIA SS2020 - Exercise Sheet 1: Introduction to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=12, error=TRUE, warning=FALSE,
        message=FALSE, 
        #echo=FALSE, results="hide", fig.show = "hide"
         echo=TRUE, results="show"
)
```

# Setup

```{r, config, echo = T}
library(ggplot2)
library(data.table)
library(magrittr)   # Needed for %>% operator
library(tidyr)
library(readxl)
library(dplyr)
```


# Introduction to ggplot

The `iris` data is included in the ggplot2 package. First load `ggplot2` package, then load `iris` data by `data(iris)`. Check `iris` data with `head(iris)`. 

1) Are there any relationships/correlations between petal length and width? How would you show it?

2) Do petal lengths and widths correlate in every species?


```{r}
## Answer: 1)
data(iris)
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point() 
```


```{r}
## Answer: 2)
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point()
```

# Histograms

Get the _gtex-annotation.csv_ data and do a histogram of the RIN (RNA integrity number) column using 10, 20, 50, 100 bins to see how this affects the visualisation.

```{r}
gtex.annotation <- fread("~/Projects/ncRNA-workshop/extdata/gtex-dummy-dataset.csv")
ggplot(gtex.annotation, aes(RIN)) + geom_histogram(bins = 50)
```


# Boxplots

Get the _gtex-annotation.csv_ data and do some boxplots of the RIN (RNA integrity number) column against the age groups. Do you see something interesting? Do the same using violin plots. Now try to combine the violin and the boxplot into one plot (use `width = 0.2`).

```{r}
ggplot(gtex.annotation, aes(age.group, RIN)) + geom_boxplot()
ggplot(gtex.annotation, aes(age.group, RIN)) + geom_violin() + geom_boxplot(width=0.2)
```

# Scatterplot

Make a scatterplot between the fake.age and the RIN for heart. Do you see any associasion between fake.age and RIN? Now color the points by `sex` so that you have the labels `Male` and `Female` on the legend. Do you see any associasion between fake.age and RIN when it is controled for sex?

```{r}
sex <- gtex.annotation$sex

sex[sex == 2 & !is.na(sex)] <- "Female"

sex[sex == 1 & !is.na(sex)] <- "Male"

gtex.annotation$sex <- sex

ggplot(gtex.annotation[tissue == "Heart"], aes(fake.age, RIN)) + geom_point()
ggplot(gtex.annotation[tissue == "Heart"], aes(fake.age, RIN, color=sex)) + geom_point()
```


# Understanding a messy dataset


The following file describes the number of times a person bought a product "a" and "b"

```{r}
messy_file <- file.path('extdata', 'example_product_data.csv')
messy_dt <- fread(messy_file)
messy_dt
```

Why is this data-set messy? Which columns should a tidy version of this table have?

```{r}
# Answer: 
# Vales are stored as column names. 
# Tidy data columns: name, product, n
```

# Fixing a messy dataset

Read the weather dataset `weather.txt`. It contains the minimal and maximal temperature on a certain city (id) over different dates (year, month, d1-d31). Why is this dataset messy? How would a tidy version of it look like? Create its tidy version.

```{r}
messy_dt <- fread("extdata/weather.txt")
messy_dt %>% head
dim(messy_dt)
```

```{r}
## Why is it messy?

## Answer: 
## 1. Variables are stored as columns (days)
## 2. A single entity is scattered across many cells (date)
## 3. Element column is not a variable.
##
## Tidy version: id, date, tmin, tmax
```


```{r}
## Fix a messy data

### First melt the table: wide -> long
dt <- melt(data = messy_dt, 
           id.vars = c("id", "year", "month", "element"), 
           variable.name = "day") 

# You can ignore the warning message
# measure.vars is missing. When missing, measure.vars will become all columns outside id.vars.
# value.name: name for the molten data values column(s). The default name is 'value'. 


### Then make the column day into integer
dt[, day := as.integer(gsub(pattern = "d", replacement = "", x = day))]


### Join all date related columns into one. Use unite or paste
# 1. Using unite():
dt <- unite(dt, "date", c("year", "month", "day"), sep = "-", remove = TRUE)

## 2. Using paste():
# dt[, date := paste(year, month, day, sep = "-")] # convert to date
# dt[, c("year", "month", "day") := NULL] # remove reduntant columns


### Dcast the table: long -> wide
dt <- dcast(data = dt, formula = ... ~ element, value.var = "value") 


### Remove entries with both NA values,
tidy_dt <- dt[!(is.na(TMAX) & is.na(TMIN))] 


## na.omit(dt) would also do the job
# tidy_dt <- na.omit(dt)

head(tidy_dt)

dim(tidy_dt)
```

```{r}
# An alternative tidy code version
tidy_dt <- messy_dt %>% 
  melt(id.vars=c('id', 'year', 'month', 'element'), na.rm=TRUE) %>% 
  .[, variable := gsub('d', '', variable)] %>% 
  unite(date, year, month, variable, sep='-') %>% 
  dcast(... ~ element) %>% 
  .[, date := as.Date(date)]
```
