---
title: "Exercise sheet: Day 1"
author: "Vangelis Theodorakis, Fatemeh Behjati, Julien Gagneur, Marcel Schulz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{CeDoSIA SS2020 - Exercise Sheet 1: Introduction to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=12, error=TRUE, warning=FALSE,
        message=FALSE, 
        echo=FALSE, results="hide"
         #echo=TRUE, results="show"
)
```

# Vectors 

First, create three named numeric vectors of size 10, 11 and 12 respectively in the following manner: 

1) One vector with the "colon" approach: *from:to*
2) One vector with the `seq()` function: *seq(from, to)*
3) And one vector with the `seq()` function and the `by` argument: *seq(from, to, by)*

For easier naming you can use the vector `letters` or `LETTERS` which contain the latin alphabet in small and capital, respectively. In order to select specific letters just use e.g. `letters[1:4]` to get the first four letters. 
Check their types. What is the outcome? Where do you think the difference comes from?


```{r}
# Answer :

# A. Create vectors
vector.1 <- 1:10
names(vector.1) <- letters[vector.1]

vector.2 <- seq(1, 11)
names(vector.2) <- letters[vector.2]

vector.3 <- seq(1, 12, by = 1)
names(vector.3) <- letters[vector.3]

typeof(vector.1)
typeof(vector.2)
typeof(vector.3)

```

# Factors
1) Create a character vector consisting of three annotations *Mutant-1, Mutant-2, Control*.
2) Using this annotation vector, create a factor where each annotation is repeated 4 times in a sequential manner (*Mutant-1, Mutant-2, Control, Mutant-1, Mutant-2, Control, ...*). In addition, the levels are the sorted annotation values.
3) Print the results.


```{r }

# Answer :
#1)
annotation <- c("Mutant-1", "Mutant-2", "Control")
#2)
test.factor <- factor(rep(annotation, 4), levels = sort(annotation))
#3)
print(test.factor)
```

# Data tables

The purpose of this exercise is to get familiarized with data.table and try out some of its useful features.

## Basic operations
Please follow the steps listed below:

1) load the library called *dslabs*

2) Access the database called *brexit_polls*. You can take a look at the the *help* documentation of this database (*?brexit_polls*) to learn about its content.

For example:

column name | Description
------------ | --------------
pollster | Pollster conducting the poll.
poll_type | Online or telephone poll.
samplesize | Sample size of poll.
remain | Proportion voting Remain.
leave | Proportion voting Leave.

3) Inspect this data by checking properties such as the class type, the number of rows and columns, its column names, the unique values in the *poll_type* column.

4) Create a new variable called *brexit_DT* and assign the data.table converted version of *brexit_polls*.

```{r}
# Answer :
library(data.table)
library(dslabs)

print("class of brexit_polls is")
class(brexit_polls)

print("dim of brexit_polls is")
dim(brexit_polls)

print("column names of brexit_polls are")
colnames(brexit_polls)

print("a small subset of data looks like")
brexit_polls[1:3, 1:5]
      
print("tissue types in data:")
unique(brexit_polls$poll_type)

brexit_DT <- as.data.table(brexit_polls)
```

## More exciting operations
Continue from the previous part and perform the following actions:


5) From *brexit_DT* get the counts of Online and Telephone polls

6) What are the mean and median values of the *samplesize*

7) Add a new column *remain_polls* to *brexit_DT* that holds the multiplication of *samplesize* to *remain*

8) What is the range of values in this newly created column?

9) How do the mean values of *undecided* look like when grouped by *pollster*? How do they look like when grouped by *poll_type*? What is this mean value when *pollster* is *YouGov*?

10) Remove the column *remain_polls* created in step 7.

```{r}
# Answer :
#5
brexit_DT[, .N, by= poll_type]

#6
brexit_DT[, .(mean_samplesize= mean(samplesize), median_samplesize= median(samplesize), sd_samplesize= sd(samplesize))]

#7
brexit_DT[, remain_polls:= samplesize * remain]

#8
brexit_DT[, range(remain_polls)]

#9
brexit_DT[, mean(undecided), by= pollster]
brexit_DT[, mean(undecided), by= poll_type]
# YouGov
brexit_DT[pollster == "YouGov", mean(undecided), by= poll_type]

#10
brexit_DT[, remain_polls := NULL]
```
