---
title: "Exercise & solution sheet: Day 1"
author: "Vangelis Theodorakis, Fatemeh Behjati, Julien Gagneur, Marcel Schulz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{CeDoSIA SS2020 - Exercise Sheet 1: Introduction to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::pdf_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=12, error=TRUE, warning=FALSE,
        message=FALSE, 
        #echo=FALSE, results="hide"
        echo=TRUE, results="show"
)
```

# Vectors 

First, create three named numeric vectors of size 10, 11 and 12 respectively in the following manner: 

1) One vector with the "colon" approach: *from:to*
2) One vector with the `seq()` function: *seq(from, to)*
3) And one vector with the `seq()` function and the `by` argument: *seq(from, to, by)*

For easier naming you can use the vector `letters` or `LETTERS` which contain the latin alphabet in small and capital, respectively. In order to select specific letters just use e.g. `letters[1:4]` to get the first four letters. 
Check their types. What is the outcome? Where do you think the difference comes from?


```{r}
# Answer :

# A. Create vectors
vector.1 <- 1:10
names(vector.1) <- letters[vector.1]

vector.2 <- seq(1, 11)
names(vector.2) <- letters[vector.2]

vector.3 <- seq(1, 12, by = 1)
names(vector.3) <- letters[vector.3]

typeof(vector.1)
typeof(vector.2)
typeof(vector.3)

```

# Factors
1) Create a character vector consisting of three annotations *Mutant-1, Mutant-2, Control*.
2) Using this annotation vector, create a factor where each annotation is repeated 4 times in a sequential manner (*Mutant-1, Mutant-2, Control, Mutant-1, Mutant-2, Control, ...*). In addition, the levels are the sorted annotation values.
3) Print the results.


```{r }

# Answer :
#1)
annotation <- c("Mutant-1", "Mutant-2", "Control")
#2)
test.factor <- factor(rep(annotation, 4), levels = sort(annotation))
#3)
print(test.factor)
```

# Data tables

The purpose of this exercise is to get familiarized with data.table and try out some of its useful features.

## Basic operations
Please follow the steps listed below:

1) Download the GTEx data (annotation v7) from the Google drive:

2) Read the file downloaded above and store it in a variable named: *data*.

3) Inspect *data* by checking properties such as the class type, the number of rows and columns, its column names, the unique values in the *SMTS* column.

If the column names are unfamiliar to you, please take a look at the description list below for some of the relevant columns to this exercise:

column name | Description
------------ | --------------
SMTS | Tissue Type (area of retrieval)
SMAFRZE | Samples in GTEx Analysis Freeze
SME1MPRT | End 1 Mapping Rate
SMEXPEFF | Expression Profiling Efficiency
```{r}
# Answer :
library(data.table)
data <- fread("~/GTEx_v7_Annotations_SampleAttributesDS.txt")

print("class of data is")
class(data)

print("dim of data is")
dim(data)

print("column names of data are")
colnames(data)

print("a small subset of data looks like")
data[1:3, 1:5]
      
print("tissue types in data:")
unique(data$SMTS)
```

## More exciting operations
Continue from the previous part and perform the following actions:

4) Subset the data based on the *Brain* cell type sample and store the result in a variable called: *data_Brain*.

5) Inspect the *data_Brain* similar to the point 3 above.

6) Examine the range of values in *SMEXPEFF* (Expression Profiling Efficiency) column of *data_Brain*. How can you make it more meaningful?

7) For *data_Brain*, compute the average of the values stored in the *SMEXPEFF* column. Also, compute the min of values stored in *SME1MPRT* (End 1 Mapping Rate).

8) Compute the correlation between the two columns mentioned above.

9) Remove the rows that are NA from *data_Brain$SMEXPEFF* and *data_Brain$SME1MPRT*. Retry the correlation on the NA-removed *data_Brain_noNA*.

**Hint:** Use the *is.na()* function to find the rows that are NA.
```{r}
# Answer :
#3
data_Brain <- data[data$SMTS == "Brain", ]

#4
print("class of data_Brain is")
class(data_Brain)

print("dim of data_Brain is")
dim(data_Brain)

print("column names of data_Brain are")
colnames(data_Brain)

print("a small subset of data_Brain looks like")
data_Brain[1:3, 1:5]
      
print("tissue types in data_Brain:e")
unique(data_Brain$SMTS)

#5
print("range of values in data_Brain$SMEXPEFF (Expression Profiling Efficiency):")
range(data_Brain$SMEXPEFF)

print("range of values in data_Brain$SMEXPEFF when NA's are removed:")
range(data_Brain$SMEXPEFF, na.rm= T)
#6
mean(data_Brain$SMEXPEFF)
mean(data_Brain$SMEXPEFF, na.rm= T)

min(data_Brain$SME1MPRT)
min(data_Brain$SME1MPRT, na.rm= T)
#7
cor(data_Brain$SME1MPRT, data_Brain$SMEXPEFF)
#8
data_Brain <- data_Brain[!is.na(data_Brain$SMEXPEFF) | !is.na(data_Brain$SME1MPRT), ]
cor(data_Brain$SME1MPRT, data_Brain$SMEXPEFF)
```

# Looping
1) Initialize a variable called *counter* by 0.
2) Using a for loop that iterates 10 times,
- create a random number drawn from a uniform distribution with *min=0* and *max=5*.
- whenever this random number is bigger than or equal to 1, increment *counter* by 1.
3) Print the final value in *counter*.

```{r}
#Answer :

# Looping
counter <- 0
for(i in seq(10)){
  r <- runif(1, 0, 5)
  if(r >= 1){
    print(r)
    counter <- counter + 1
  }
}
print(counter)
```
# Functions
1) Write a function named *get_counts* that takes a GTEx data table as input and outputs the total counts of rows that the sample tissue type (*SMTS*) is *Heart* and the sample analysis freeze (*SMAFRZE*) is *RNASEQ*.
2) How about if you try the same but for *Blood*.

- If this task was too easy, can you modify your function such that instead of taking only one argument, it takes two additional ones, one for the *SMTS* and another for *SMAFRZE*. Iterate over all possible values of *SMTS* (**Hint:** *unique(data$SMTS)*) and call your function by providing the sample tissue type.

```{r}
#Answer :
get_counts <- function(gtex){
  counter <- 0
  for(i in seq(nrow(gtex)))
    if(gtex$SMTS[i] == "Heart" & gtex$SMAFRZE[i] == "RNASEQ"){
      counter <- counter + 1
    }
  return(counter)
}

fun_res <- get_counts(data)
print(fun_res)

#modified version
get_counts2 <- function(gtex, var1, var2){
  counter <- 0
  for(i in seq(nrow(gtex)))
    if(gtex$SMTS[i] == var1 & gtex$SMAFRZE[i] == var2){
      counter <- counter + 1
    }
  return(counter)
}

all_tissues <- unique(data$SMTS)
for(ts in all_tissues){
  fun_res <- get_counts2(data, ts, "RNASEQ")
  print(paste("Number of RNASEQ cases for", ts, ":", fun_res))
}

```
# R Markdown
Downloaded and stored the *sample_annotation.tsv* file from Google drive. Then, create an Rmarkdown file and perform the following tasks:
1) Read the *sample_annotation.tsv* file.
2) Create a new variable containing the counts of each *tissue* existing in the data.
3) Use the *barplot* function to plot the number of tissue types in the GTEx data.
4) Try to sort the bars according to the tissue counts (optional).
``` {r}
gtex.annotation <- fread("../../extdata/sample_annotation.tsv")
samples.per.tissue <- gtex.annotation[, .N, by = "tissue"]
barplot(N~tissue, samples.per.tissue, las = 2, xlab = "",
        main= "Number of samples per tissue")
samples.per.tissue.ordered <- samples.per.tissue[order(-N)]
samples.per.tissue.ordered$tissue <- factor(samples.per.tissue.ordered$tissue,
                                            levels = samples.per.tissue.ordered$tissue)
barplot(N~tissue, samples.per.tissue.ordered, las = 2, xlab = "",
        main= "Number of samples per tissue (sorted)")
```