# Data Wrangling using Data.table

```{r include=F}
DATADIR <- file.path('../extdata')
```

Data Wrangling includes tasks of processing raw data into interesting formats....


Data.tables are a modern implementation of tables in R. We will exclusively used data.tables in this course. Base R provides a similar structure called data.frames.
However, those are a lot slower and often a little more complicated to use, so we highly recommend using data.table.

Similar to a data.frame, but because it modifies columns by *reference*, a data.table is a memory efficient and faster implementation of data.frame offering:

* sub-setting
* ordering
* merging
 
It accepts all data.frame functions, it has a shorter and more flexible syntax (not so straightforward in the beginning but pays off) and saves time on two fronts:

* programming (easier to code, read, debug and maintain)
* computing (fast and memory efficient)

The general form of data.table syntax is:

    DT[ i,  j,  by ] # + extra arguments
        |   |   |
        |   |    -------> grouped by what?
        |    -------> what to do?
         ---> on which rows?
         
The way to read this out loud is: "Take DT, subset rows by i, then compute j grouped by by. Here are some basic usage examples expanding on this definition.

Like data.frame, data.table is a list of vectors. It doesn't have row names; instead, it's a collection of attributes.

Each column can be a different type, including a list

It has enhanced functionality in `[`. We can make operations inside `[]`.

For a detailed documentation go check the help page for data.table!

## Creating and loading data 

To create a data.table, we just name its columns and populate them. All the columns have to have the same length.

```{r}
# install.packages("data.table")
library(data.table)
DT <- data.table(x = rep(c("a","b","c"), each = 3), y = c(1, 3, 6), v = 1:9)
DT # note how column y was recycled
```

If we want to convert any other R object to a data.table, all we have to do is to call the `as.data.table()` function.

```{r}
# This way we can for example convert any inbuilt data set into a data.table:
titanic_dt <- as.data.table(Titanic)
class(titanic_dt)
```

Here you can see that the class function informs us that DT is both a data.table and a data.frame as data.tables inherit
from data.frames, an older table class in R.

Alternativly we can read files from disk and process them using data.table. 
The easiest way to do so is to use the function `fread()`.

```{r, eval=FALSE}
flights <- fread('path_to_file/flightsLAX.csv')
```

```{r, echo=FALSE}
flights <- fread(file.path(DATADIR, 'flightsLAX.csv'))
flights <- flights[, .(YEAR, MONTH, DAY, DAY_OF_WEEK, AIRLINE, FLIGHT_NUMBER, TAIL_NUMBER, ORIGIN_AIRPORT,
                        DESTINATION_AIRPORT, DEPARTURE_TIME, AIR_TIME, DISTANCE, ARRIVAL_TIME)]
```
Calling `head()` on the table showes us the first few lines of the table and we observe that reading the file was succesfull.
```{r}
head(flights)
```

## Inspecting tables

A first step in any analysis should involve inspecting the data we just read in. 
This often starts by looking the head of the table as we did above.

The next information we are often interested in is the size of our data set.
We can use the following commands to obtain it.
```{r}
nrow(flights)   # ncol(flights)
dim(flights)
```

### Technical detail

The structure of the data.table essentially is a list of list. To get the dimensions of our data.table we can either use the `dim()` function to get the dimensions of our data.table (M rows x N columns) or to use the `ncol()` `nrow()` functions.

```{r}
class(DT)
is.list(DT)
nrow(DT)   # ncol(DT)
dim(DT)
```


Next we are often interested in the range or the number of occurence of values in certain columns. 
To obtain this information we can use the summary command on the table.

```{r}
summary(flights[,1:6])

```

This provides us already a lot of information about our data. We can for example see that all data is from 2015 as all values in the YEAR column are 2015.
But for categorical data this is not very insight full, as we can see for the AIRLINE column.

To investigate categorical columns we can have a look at their unique elements using:
```{r}
flights[, unique(AIRLINE)]

```
This command provided us the airline identifiers present in the dataset.
Another valueable information for categoricals is how often each category occurs. This can be obtained using the following commands:
```{r}
flights[, table(AIRLINE)]
```
Or:

```{r}

flights[, .N, by = 'AIRLINE']

```

Especially the second command will be explained in detail later.

```{r}



```


## Accessing rows and columns of data.tables

Data.tables can be subsetted by indices, if we for example care about the second element of the table.
We can do the following.
```{r}
flights[2, ]   # Access the 2nd row (also flights[2] or flights[i = 2])
flights[1:2]   # Access multiple consecutive rows.
flights[c(3, 5)]  # Access multiple rows.
```
Additionally we can do the same for columns:
```{r}
head(flights[, 2 ])
head(flights[, c(3, 5)])
```
Alternatively to subsetting by indices we can subset columns by their names.

```{r}
head(flights[, .(MONTH)])
head(flights[, .(DAY, AIRLINE)])
```

## Sub-setting rows according to some condition

A often more usefull way is to subset rows using conditions!
We can subset a data table using the following operators:

* `==`
* `<` 
* `>`
* `!=`
* `%in%`

For example if we are interested in analysing all the flights operated by "AA" (American Airlines) we can do that using the following command:

```{r}
head(flights[AIRLINE == "AA"])
```
Alternatively if we are interested in all flights from any destination to the airports in NYC (JFK and LGA), we can subset the rows using the following command:
```{r}
flights[DESTINATION_AIRPORT %in% c("LGA", "JFK")]
```

Additionally we can concatenate multiple condition using the logical OR `|` or AND `&` operator.
If we for example want to inspect all flights departing between 6am and 7am operated by American Airlines we can use the following statement:
```{r}
head(flights[AIRLINE == "AA" & DEPARTURE_TIME > 600 & DEPARTURE_TIME < 700])
```


## Accessing a data.table by columns (DT[i, **j**, by])

Once you're inside the ``[]``, you're in the data.table environment. Inside this scope, there's no need to put the column names in quotation marks, as columns are seen as variables already.

It is not advisable to access a column by its number. **Use the column name instead**.

The table format can change (new columns can be added), so using column names prevents bugs e.g. if you have a data set with 50 columns, how do you know which one is column 18?

This way the code is more readable: `flights[, TAIL_NUMBER]` instead of `flights[, 7]`.

```{r}
flights[1:10, TAIL_NUMBER]    # Access column x (also DT$x or DT[j=x]). 
flights[4, TAIL_NUMBER]   # Access a specific cell.
```

When accessing many columns, we probably want to return a data.table instead of a vector. For that, we need to provide R with a list, so we use ``list(colA, colB)``, or its simplified version ``.(colA, colB)``.

```{r}
# Note that 1 and 3 were coerced into strings because a vector can have only 1 type
flights[1:2, c(TAIL_NUMBER, ORIGIN_AIRPORT)]   

# Access a specific subset. Data.frame: DF[1:2, c("x","y")] 
flights[1:2, list(TAIL_NUMBER, ORIGIN_AIRPORT)]

# Same as before.
flights[1:2, .(TAIL_NUMBER, ORIGIN_AIRPORT)]
```

### Technical Detail Data.table environment

The following examples are to show how the ``[]`` bring us inside the data.table environment.

Using DT and DF, compute the product of columns *y* and *v*.

Use the ``with(data, expr, ...)`` function, which evaluates an R expression in an environment constructed from ``data``.

```{r eval = FALSE}
# We enter the environment of DT, and simply compute the product
with(DT, y*v)

# Easier way, with [] we're inside the environment. Data.table 
# runs a with(dt,...) on the j argument
DT[, y*v]

# In data.frame, the [] doesn't imply we're inside the environment
DF[, y*v] 
```


## Basic operations

We saw already that inside the ``[]``, columns are seen as variables, so we can apply functions to them.
```{r}
# Similar to mean(flights[, AIR_TIME])
flights[, mean(AIR_TIME, na.rm=TRUE)]   
flights[AIRLINE == "OO", mean(AIR_TIME, na.rm=TRUE)]
```

To compute operations in multiple columns, we must provide a list (unless we want the result to be a vector).

```{r}
# Same as flights[, .(mean(AIR_TIME), median(AIR_TIME))]
flights[, list(mean(AIR_TIME, na.rm=TRUE), median(AIR_TIME, na.rm=TRUE))]

# Give meaningful names
flights[, .(mean_AIR_TIME = mean(AIR_TIME, na.rm=TRUE), median_AIR_TIME = median(AIR_TIME, na.rm=TRUE))] 
```

### Technical detail We can also apply basic operations on multiple columns

``sapply(DT, FUN)``, applies function FUN column-wise to DT. Remember that ``sapply`` returns a vector, while ``lapply`` returns a list.

```{r, linewidth=80, error=TRUE}
sapply(iris_dt, class)   # Try the same with lapply
sapply(iris_dt, sum)
```

```{r}

# Note that we can access columns stored as variables by setting with=F.
# In this case, `colnames(iris_dt)!="Species"` returns a logical vector and `
# iris_dt` is subseted by the logical vector

# Same as sapply(iris_dt[, 1:4], sum)
#sapply(iris_dt[, colnames(iris_dt)!="Species", with = F], sum) 
```

## The 'by' option (DT[i, j, **by**])

The ``by = `` option allows us to apply a function to groups wtihin a data.table. For example, we can use the ``by = `` to compute the mean flight time per airline:

```{r}
# Compute the mean air time of every airline
flights[, .(mean_AIRTIME = mean(AIR_TIME, na.rm=TRUE)), by = AIRLINE]
```

This way we can easily spot that one airline conducts on average shorter flights.

```{r}
# Compute the mean and standard deviation of the air time of every airline
flights[, .(mean_AIRTIME = mean(AIR_TIME, na.rm=TRUE), sd_AIR_TIME = sd(AIR_TIME, na.rm=TRUE)), by = AIRLINE]
```

Although we could write ``flights[i = 5, j = AIRLINE]``, we usually ommit the ``i =`` and ``j =`` from the syntax, and write ``flights[5, ARILINE]`` instead. However, for clarity we usually include the ``by =`` in the syntax.

## Counting occurences (the `.N` command)

The ``.N`` is a special in-built variable that counts the number observations within a table.

Evaluating ``.N`` alone is equal to `nrow()` of a table.
```{r}
flights[, .N]
nrow(flights)
```

But the ``.N`` command becomes a lot more powerful when used with grouping or conditioning.
We already saw earlier how we can use it to count the number of occurrences of elements in categorical columns.

```{r}
# Get the number of flights for each AIRLINE
flights[, .N, by = 'AIRLINE']
```

Remembering the data.table definition: "Take **DT**, subset rows using **i**, then select or calculate **j**, grouped by **by**",
we can build even more powerful statements using all three elements.
```{r}
# For each airline, get the number of observations to NYC (JFK)
flights[DESTINATION_AIRPORT == "JFK", .N, by = 'AIRLINE']
```

## Creating new columns (the := command)

The ``:=`` operator updates the data.table you are working on, so writing DT <- DT[,... := ...] is redundant. This operator, plus all ``set`` functions (TODO clarify), change their input by *reference*. No copy of the object is made, that is why it is faster and uses less memory.

```{r}
# Add a new column called SPEED whose value is the DISTANCE divided by AIR_TIME
flights[, SPEED := DISTANCE / AIR_TIME * 60]
head(flights)
```

Having computed a new column using the ``:=`` operator we can use it for further analyses.
```{r}
flights[, .(mean_AIR_TIME = mean(AIR_TIME, na.rm=TRUE), 
            mean_SPEED = mean(SPEED, na.rm=TRUE),
            mean_DISTANCE = mean(DISTANCE, na.rm=TRUE)
            ), by=AIRLINE] 
```
Now we can see that the flights by the carrier "OO" are not just shorter, but also slow. 
This could for example lead us to the hypothesis, that "OO" is a small regional carrier, which operates slower planes.

Additionally we can use the ``:=`` operator to remove columns. If we for example observe that tail numbers are not important for our analysis we can remove them with the following statement:
```{r}
flights[, TAIL_NUMBER := NULL]
head(flights)
```
Here we observe, that the tail numbers are gone from the data.table

### Technical Detail Multiple assignments

With the following syntax we can assign multiple new columns at once. 

```{r}
# Add columns with sepal and petal area. Note the syntax of multiple assignment.
#iris_dt[, `:=` (Sepal.Area = Sepal.Length * Sepal.Width,
#                Petal.Area = Petal.Length * Petal.Width)][1:3]
```

You can also delete columns by using the ``:=`` command.

```{r}

# Let's assume setosa flowers are orange, versicolor purple and virginica pink.
# Add a column with these colors.
#iris_dt[Species == "setosa", color := "orange"] 
#iris_dt[Species == "versicolor", color := "purple"]
#iris_dt[Species == "virginica", color := "pink"]
#unique(iris_dt[, .(Species, color)])

# We can delete this new column by setting it to NULL
#iris_dt[, color := NULL]
#colnames(iris_dt)
```

## By reference

What do we mean when we say that data.table modifies columns *by reference*? It means that no new copy of the object is made in the memory, unless we actually create one using ``copy()``.
```{r}
or_dt <- data.table(a = 1:10, b = 11:20)
# No new object is created, both new_dt and or_dt point to the same memory chunk.
new_dt <- or_dt   
new_dt[, ab := a*b]
colnames(or_dt)   # or_dt was also affected by changes in new_dt

or_dt <- data.table(a = 1:10, b = 11:20)
copy_dt <- copy(or_dt)   # By creating a copy, we have 2 objects in memory
copy_dt[, ab := a*b]
colnames(or_dt)    # Changes in the copy don't affect the original
```


  
## Merging

Merge, join or combine 2 data tables into one by common column(s).

``merge(x, y, by, by.x, by.y, all, all.x, all.y, ...)``

In data.table, the default is to merge by shared *key* columns while in data.frame, the default is to merge by columns with the same name. We can also specify the columns to merge ``by``.

There are different types of merging:

* **Inner (default)**: consider only rows with matching values in the key columns.
* **Outer or full**: return all rows and columns from x and y. If there are no matching values, return NAs.
* **Left (all.x)**: consider all rows from x, even if they have no matching row in y.
* **Right (all.y)**: consider all rows from y, even if they have no matching row in x.

### Merging - example

If we for example want to know how the airports are called behind the IATA_CODE codes, we can load another tabe containing additional information about the airports.
```{r, eval=FALSE}
airports <- fread('path_to_file/airports.csv')
```

```{r, echo=FALSE}
airports <-  fread(file.path(DATADIR, 'airports.csv'))
```
```{r}
head(airports)
```

```{r}
head(merge(flights, airports, by.x = "DESTINATION_AIRPORT", by.y = 'IATA_CODE'))
```

### Merging - inner and outer examples

Below we created some artificial tables to showcase the differences between the different types of merges.
```{r}
dt1 <- data.table(table = "table1", id = c(1,2,3), a = rnorm(3))
dt2 <- data.table(table = "table2", id = c(1,2,4), b = rnorm(3))
```

```{r}
# Inner merge: default one, all = FALSE
merge(dt1, dt2, by = "id", all = F)
```

```{r}
# Outer (full) merge: all = TRUE
merge(dt1, dt2, by = "id", all = T)
```

Note that the column order got changed after the merging.
 
```{r}
# Left merge: all.x = TRUE
merge(dt1, dt2, by = "id", all.x = T)[]
```
 
```{r}
# Right merge: all.y = TRUE
merge(dt1, dt2, by = "id", all.y = T)[]
```

## Summary

By now, you should be able to answer the following questions:

* Why do we say that data.table is an enhanced data.frame?
* How to subset by rows or columns? Remember: DT[i, j, by].
* How to add columns?
* How to make operations with different columns?
* Which are the different types of merging?

Even if you were able to answer them, practice:

* Check all vignettes and reference manual from <https://cran.r-project.org/web/packages/data.table/>
* A really concise cheat sheet: 
* <https://s3.amazonaws.com/../assets.datacamp.com/img/blog/data+table+cheat+sheet.pdf>

## Data.table resources

<https://cran.r-project.org/web/packages/data.table/>

<https://s3.amazonaws.com/../assets.datacamp.com/img/blog/data+table+cheat+sheet.pdf>

<http://r4ds.had.co.nz/relational-data.html>

<http://adv-r.had.co.nz/Environments.html>

\pagebreak
