```{r include=FALSE}
options("width"=80)
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, cache=T, 
    results="show",
    out.width="400px", out.height="250px",
    fig.align="center",
    dpi=200, tidy = TRUE, tidy.opts=list(width.cutoff=80))

error_hook <- knitr::knit_hooks$get("error")
knitr::knit_hooks$set(
  error = function(x, options) {
    if (!is.null(n <- options$linewidth)) {
      x = knitr:::split_lines(x)
      if (any(nchar(x) > n))
        x = strwrap(x, width = n)
      x = paste(x, collapse = '\n')
    }
    error_hook(x, options)
  }
)
```

## Assignments

Assignments in R are preferably of this form:

```{r eval=F}
objectName <- value
```

It is also possible to assign with the *equal* sign.

```{r eval=F}
objectName = value
```

```{r}
x <- 5 # Both methods have the same outcome
y = 5
```

```{r}
x
y
```

However, the "equal" sign is used for argument passing to functions. Thus if nesting, the equal sign will be interpreted as a argument assignment and might throw an error:

```{r, linewidth=80, error=TRUE}
## We want to measure the running time of the inner product of a large vector and 
## assign the outcome of the function to a variable simultaneously
system.time(a <- t(1:1e6)%*%(1:1e6))
system.time(a = t(1:1e6)%*%(1:1e6)) ## This would cause an error
```

Right Alt + - gives a quick shortcut to add **<-** ;)


## Vectors

Vectors are 1-dimensional data structures which can contain one or more variables, regardless of their type.

Usually created with `c()` (from *concatenation*):
```{r}
c(1, 5, 8, 10)
str(c(1, 5, 8, 10))
length(c(1, 5, 8, 10))
c("a", "B", "cc")
c(TRUE, FALSE, c(TRUE, TRUE))
c(1, "B", FALSE)
```

There are multiple ways to create a numeric sequence depending on the desired result.

Usually for an integer sequence *from:to* is enough. For different results we may need to use `seq()` function utilizing its arguments to increase by a *step* of our choice or to split the range depending on the desired length of the output.

```{r}
1:10 
seq(from = 1, to = 10, by = .3)
seq(from = 1, to = 10, length.out = 12)
```

Keep in mind that sometimes different approaches, give different data types as results so when weird things happen, always check the documentation (devil is in the details)!

For non numeric values e.g. logical, character, `rep()` function comes also in handy. 

The *replicate* function `rep()` replicates a vector a certain number of *times* and concatenates them: 
```{r}
rep(c(TRUE, FALSE), times = 5)
```

To replicate *each* entry of the input vector at the time:
```{r}
rep(c(TRUE, FALSE), each = 5)
```

Vector elements are accessed via the ```[``` operator:
```{r}
## Create an A,B,C,D,E vector
x <- LETTERS[1:5]
x
## access the third entry
x[3]
## modify the third entry
x[3] <- 'Z'
x
```

Elements in vectors can have names. 
Using names instead of index to access entries in a vector make code more robust to re-ordering, sub-setting, and changes in data input.  
Names can be created at initialization or set afterwards with `names()`. 

```{r}
x <- c(a = 1, b = 2, c = 3)
x
names(x) <- c("A", "B", "C")
x
```

Names don't have to be unique, but should preferably be, as sub-setting by names will only return the first match
```{r}
x <- c(a = 1, a = 2, b = 3)
x["a"]
```

Not all elements need to have names.
```{r}
c(a = 1, 2, 3)
```

## Factors

```{r}
x <- factor(c("red", "yellow", "red","green", "green"))
x
```

* A factor is a is used to store categorical data. The distinct categories are called 'levels'.

* They belong to a special `class()`, *factor*, which makes them behave differently from regular integer vectors.

* Factors cannot be altered in the same way as vectors. Only their levels can be altered.

* Factors are built on top of integer vectors: the values are integer indexes in the dictionary of levels.

* They occupy less space in memory than characters, since they are stored as integers.


Level | Integer 
------|--------
green  | 1 
red | 2 
yellow | 3 

* Factors are typically constructed with `factor()`. By default the levels are the unique values, sorted by alphanumerical order. 
```{r}
x <- factor(c("red", "yellow", "red","green", "green"))
x
```

`levels()` gives the levels in ascending order of the underlying encoding integers.
```{r}
levels(x)
```

The order of the levels can be forced:
```{r}
x <- factor(c("red", "yellow", "red","green", "green"), levels=c("green", "yellow", "red"))
x
levels(x)
```

The order of the levels is then used for all function requiring comparisons, e.g. sorting 
```{r}
sort(x)
```

Only level values can be used:
```{r}
x <- factor(c("red", "yellow", "red","green", "green"))
x[2] <- "blue"
x
```

**Be aware** that R does not prevent you from combining factors:

Do not try to combine factors, especially if levels are not the same!

```{r}
c(x, factor("blue"))
```

# Data Wrangling using Data.table

```{r include=F}
DATADIR <- file.path('extdata')
```

Data Wrangling includes tasks of processing raw data into interesting formats....

Data.tables are a modern implementation of tables in R. We will exclusively used data.tables in this course. Base R provides a similar structure called data.frames.
However, those are a lot slower and often a little more complicated to use, so we highly recommend using data.table.

Similar to a data.frame, but because it modifies columns by *reference*, a data.table is a memory efficient and faster implementation of data.frame offering:

* sub-setting
* ordering
* merging
 
It accepts all data.frame functions, it has a shorter and more flexible syntax (not so straightforward in the beginning but pays off) and saves time on two fronts:

* programming (easier to code, read, debug and maintain)
* computing (fast and memory efficient)

The general form of data.table syntax is:

    DT[ i,  j,  by ] # + extra arguments
        |   |   |
        |   |    -------> grouped by what?
        |    -------> what to do?
         ---> on which rows?
         
The way to read this out loud is: "Take DT, subset rows by i, then compute j grouped by by. Here are some basic usage examples expanding on this definition.

Like data.frame, data.table is a list of vectors. It doesn't have row names; instead, it's a collection of attributes.

Each column can be a different type, including a list

It has enhanced functionality in `[`. We can make operations inside `[]`.

For a detailed documentation go check the help page for data.table!

## Creating and loading data 

To create a data.table, we just name its columns and populate them. All the columns have to have the same length.

```{r}
# install.packages("data.table")
library(data.table)
DT <- data.table(x = rep(c("a","b","c"), each = 3), y = c(1, 3, 6), v = 1:9)
DT # note how column y was recycled
```

If we want to convert any other R object to a data.table, all we have to do is to call the `as.data.table()` function.

```{r}
# This way we can for example convert any inbuilt data set into a data.table:
titanic_dt <- as.data.table(Titanic)
class(titanic_dt)
```

Here you can see that the class function informs us that DT is both a data.table and a data.frame as data.tables inherit
from data.frames, an older table class in R.

Alternativly we can read files from disk and process them using data.table. 
The easiest way to do so is to use the function `fread()`.

```{r, eval=FALSE}
flights <- fread('path_to_file/flightsLAX.csv')
```

```{r, echo=FALSE}
flights <- fread(file.path(DATADIR, '../extdata/flightsLAX.csv'))
flights <- flights[, .(YEAR, MONTH, DAY, DAY_OF_WEEK, AIRLINE, FLIGHT_NUMBER, TAIL_NUMBER, ORIGIN_AIRPORT,
                        DESTINATION_AIRPORT, DEPARTURE_TIME, AIR_TIME, DISTANCE, ARRIVAL_TIME)]
```

Calling `head()` on the table shows us the first few lines of the table and we observe that reading the file was succesfull.

```{r}
head(flights)
```

## Inspecting tables

A first step in any analysis should involve inspecting the data we just read in. 
This often starts by looking the head of the table as we did above.

The next information we are often interested in is the size of our data set.
We can use the following commands to obtain it.

```{r}
nrow(flights)   # ncol(flights)
dim(flights)
```

### Technical detail

The structure of the data.table essentially is a list of list. To get the dimensions of our data.table we can either use the `dim()` function to get the dimensions of our data.table (M rows x N columns) or to use the `ncol()` `nrow()` functions.

```{r}
class(DT)
is.list(DT)
nrow(DT)   # ncol(DT)
dim(DT)
```


Next we are often interested in the range or the number of occurence of values in certain columns. 
To obtain this information we can use the summary command on the table.

```{r}
summary(flights[,1:6])

```

This provides us already a lot of information about our data. We can for example see that all data is from 2015 as all values in the YEAR column are 2015.
But for categorical data this is not very insight full, as we can see for the AIRLINE column.

To investigate categorical columns we can have a look at their unique elements using:
```{r}
flights[, unique(AIRLINE)]

```

This command provided us the airline identifiers present in the dataset.

Another valueable information for categoricals is how often each category occurs. This can be obtained using the following commands:

```{r}
flights[, table(AIRLINE)]
```

Or:

```{r}

flights[, .N, by = 'AIRLINE']

```

Especially the second command will be explained in detail later.

## Accessing rows and columns of data.tables

Data.tables can be subsetted by indices, if we for example care about the second element of the table.
We can do the following.
```{r}
flights[2, ]   # Access the 2nd row (also flights[2] or flights[i = 2])
flights[1:2]   # Access multiple consecutive rows.
flights[c(3, 5)]  # Access multiple rows.
```

Additionally we can do the same for columns:

```{r}
head(flights[, 2 ])
head(flights[, c(3, 5)])
```

Alternatively to subsetting by indices we can subset columns by their names.

```{r}
head(flights[, .(MONTH)])
head(flights[, .(DAY, AIRLINE)])
```

## Sub-setting rows according to some condition

A often more usefull way is to subset rows using conditions!
We can subset a data table using the following operators:

* `==`
* `<` 
* `>`
* `!=`
* `%in%`

For example if we are interested in analysing all the flights operated by "AA" (American Airlines) we can do that using the following command:

```{r}
head(flights[AIRLINE == "AA"])
```

Alternatively if we are interested in all flights from any destination to the airports in NYC (JFK and LGA), we can subset the rows using the following command:

```{r}
flights[DESTINATION_AIRPORT %in% c("LGA", "JFK")]
```

Additionally we can concatenate multiple condition using the logical OR `|` or AND `&` operator.
If we for example want to inspect all flights departing between 6am and 7am operated by American Airlines we can use the following statement:

```{r}
head(flights[AIRLINE == "AA" & DEPARTURE_TIME > 600 & DEPARTURE_TIME < 700])
```

## Accessing a data.table by columns (DT[i, **j**, by])

Once you're inside the ``[]``, you're in the data.table environment. Inside this scope, there's no need to put the column names in quotation marks, as columns are seen as variables already.

It is not advisable to access a column by its number. **Use the column name instead**.

The table format can change (new columns can be added), so using column names prevents bugs e.g. if you have a data set with 50 columns, how do you know which one is column 18?

This way the code is more readable: `flights[, TAIL_NUMBER]` instead of `flights[, 7]`.

```{r}
flights[1:10, TAIL_NUMBER]    # Access column x (also DT$x or DT[j=x]). 
flights[4, TAIL_NUMBER]   # Access a specific cell.
```

When accessing many columns, we probably want to return a data.table instead of a vector. For that, we need to provide R with a list, so we use ``list(colA, colB)``, or its simplified version ``.(colA, colB)``.

```{r}
# Note that 1 and 3 were coerced into strings because a vector can have only 1 type
flights[1:2, c(TAIL_NUMBER, ORIGIN_AIRPORT)]   

# Access a specific subset. Data.frame: DF[1:2, c("x","y")] 
flights[1:2, list(TAIL_NUMBER, ORIGIN_AIRPORT)]

# Same as before.
flights[1:2, .(TAIL_NUMBER, ORIGIN_AIRPORT)]
```

## Basic operations

We saw already that inside the ``[]``, columns are seen as variables, so we can apply functions to them.

```{r}
# Similar to mean(flights[, AIR_TIME])
flights[, mean(AIR_TIME, na.rm=TRUE)]   
flights[AIRLINE == "OO", mean(AIR_TIME, na.rm=TRUE)]
```

To compute operations in multiple columns, we must provide a list (unless we want the result to be a vector).

```{r}
# Same as flights[, .(mean(AIR_TIME), median(AIR_TIME))]
flights[, list(mean(AIR_TIME, na.rm=TRUE), median(AIR_TIME, na.rm=TRUE))]

# Give meaningful names
flights[, .(mean_AIR_TIME = mean(AIR_TIME, na.rm=TRUE), median_AIR_TIME = median(AIR_TIME, na.rm=TRUE))] 
```


```{r}

# Note that we can access columns stored as variables by setting with=F.
# In this case, `colnames(iris_dt)!="Species"` returns a logical vector and `
# iris_dt` is subseted by the logical vector

# Same as sapply(iris_dt[, 1:4], sum)
#sapply(iris_dt[, colnames(iris_dt)!="Species", with = F], sum) 
```

## The 'by' option (DT[i, j, **by**])

The ``by = `` option allows us to apply a function to groups wtihin a data.table. For example, we can use the ``by = `` to compute the mean flight time per airline:

```{r}
# Compute the mean air time of every airline
flights[, .(mean_AIRTIME = mean(AIR_TIME, na.rm=TRUE)), by = AIRLINE]
```

This way we can easily spot that one airline conducts on average shorter flights.

```{r}
# Compute the mean and standard deviation of the air time of every airline
flights[, .(mean_AIRTIME = mean(AIR_TIME, na.rm=TRUE), sd_AIR_TIME = sd(AIR_TIME, na.rm=TRUE)), by = AIRLINE]
```

Although we could write ``flights[i = 5, j = AIRLINE]``, we usually ommit the ``i =`` and ``j =`` from the syntax, and write ``flights[5, ARILINE]`` instead. However, for clarity we usually include the ``by =`` in the syntax.

## Counting occurences (the `.N` command)

The ``.N`` is a special in-built variable that counts the number observations within a table.

Evaluating ``.N`` alone is equal to `nrow()` of a table.
```{r}
flights[, .N]
nrow(flights)
```

But the ``.N`` command becomes a lot more powerful when used with grouping or conditioning.
We already saw earlier how we can use it to count the number of occurrences of elements in categorical columns.

```{r}
# Get the number of flights for each AIRLINE
flights[, .N, by = 'AIRLINE']
```

Remembering the data.table definition: "Take **DT**, subset rows using **i**, then select or calculate **j**, grouped by **by**",
we can build even more powerful statements using all three elements.
```{r}
# For each airline, get the number of observations to NYC (JFK)
flights[DESTINATION_AIRPORT == "JFK", .N, by = 'AIRLINE']
```

## Creating new columns (the := command)

The ``:=`` operator updates the data.table you are working on, so writing DT <- DT[,... := ...] is redundant. This operator, plus all ``set`` functions (TODO clarify), change their input by *reference*. No copy of the object is made, that is why it is faster and uses less memory.

```{r}
# Add a new column called SPEED whose value is the DISTANCE divided by AIR_TIME
flights[, SPEED := DISTANCE / AIR_TIME * 60]
head(flights)
```

Having computed a new column using the ``:=`` operator we can use it for further analyses.

```{r}
flights[, .(mean_AIR_TIME = mean(AIR_TIME, na.rm=TRUE), 
            mean_SPEED = mean(SPEED, na.rm=TRUE),
            mean_DISTANCE = mean(DISTANCE, na.rm=TRUE)
            ), by=AIRLINE] 
```

Now we can see that the flights by the carrier "OO" are not just shorter, but also slow. 
This could for example lead us to the hypothesis, that "OO" is a small regional carrier, which operates slower planes.

Additionally we can use the ``:=`` operator to remove columns. If we for example observe that tail numbers are not important for our analysis we can remove them with the following statement:

```{r}
flights[, TAIL_NUMBER := NULL]
head(flights)
```
Here we observe, that the tail numbers are gone from the data.table

## By reference

What do we mean when we say that data.table modifies columns *by reference*? It means that no new copy of the object is made in the memory, unless we actually create one using ``copy()``.
```{r}
or_dt <- data.table(a = 1:10, b = 11:20)
# No new object is created, both new_dt and or_dt point to the same memory chunk.
new_dt <- or_dt   
new_dt[, ab := a*b]
colnames(or_dt)   # or_dt was also affected by changes in new_dt

or_dt <- data.table(a = 1:10, b = 11:20)
copy_dt <- copy(or_dt)   # By creating a copy, we have 2 objects in memory
copy_dt[, ab := a*b]
colnames(or_dt)    # Changes in the copy don't affect the original
```

## Math

R supports various mathematical and logical operations between scalars, vectors etc.

* Basic mathematical operations: `+`, `*`, `-`, `/`



## Binary comparison

- Element-wise Binary comparison return a vector of same length as the input
  - **==** element-wise equality
  - **!=** element-wise inequality
  - **<**, **<=** element-wise smaller (or equal)
  - **>**, **>=** element-wise greater (or equal)

```{r}
1:5 == 1:5
```  

* Binary comparison with a single Boolean statement as output
  * `identical(x, y)` exact equality
  * `all(x)` are all TRUE?
```{r}
identical(1:5, 1:5)
```

## Sub-setting

* There are three accessing methods
  * **[** For vectors

* Six ways of sub-setting
  * By positive integers
  * By negative integers
  * By logical vectors
  * Empty sub-setting
  * By Zero
  * Character vector

### Atomic vectors

```{r}
x <- LETTERS[1:5]
x
## By positive integers
x[c(1,3)]
## By negative integers
x[-c(1,3)]
```

```{r}
## By logical vectors
x[c(TRUE, TRUE, TRUE, FALSE, FALSE)]
## By logical vectors
x[x <= "C"]
```

`NAs` in the input vector will always produce an `NA` in the output:
```{r}
x[c(1,2,NA)]
```

We can also sub-setting by character vectors, if names are present:
```{r}
names(x) <- letters[1:5]
x[c("a", "b")]
```

### Sub-setting and assignment

All sub-setting operators can be combined with assignments to modify a subset of values.

```{r}
x <- c(a = 1, b = 2, c = 3)
x[2] <- 10
x
df <- data.frame(a = 1:2, b = 3:4)
df
df$b <- c("first", "second")
df

```

## Functions

Function are a nice way to break down a big problem (bake a pizza) to smaller, manageable sub-problems (make the dough, make the sauce, add toppings, bake the pizza). They offer code re-usage (save time for important things, like wine drinking), easier testing of functionality and cleaner code!

The syntax to write your own functions is the following:

```{r, eval=FALSE}
myFunction <- function(args1, args2 = default, ...) {
  body
  return(result)
}
```

* If the `return` statement is missing, the result of the last line will be returned
* The parameters can be assigned default values
* If the default value is a vector, the first value will be used by default
* The *three dots* (`...`) in the function arguments are called **ellipsis** and are used as a placeholder for any arguments passed to a function call inside. This is useful when you don't want to push all parameters from embodied function to your functions or allow for generic functions without pre-specifying argument names.

```{r, eval=FALSE}
myFunction <- function(args1, args2 = c("option1", "option2"), ...) {
  result <- doSomething(args1, ...)
  return(result)
}
myFunction(args1, args2, param1)
```


## Conditional Statements

A typical way to control the flow of the execution of your code is to have conditional statements.

The typical setup is:

```{r, eval=FALSE}
if(statement_to_test) {
  print("Statement is true")
} else {
  print("Statement is false")
}
```

Long version:
```{r}
if(1 < 2) {
  print("TRUE")
} else {
  print("FALSE")
}
```


Short version if the To-Do step fits into one line:
```{r}
{
if(1 > 2) print("TRUE")
else print("FALSE")
}
```

The curly brackets around the if-else statement are only needed in a script to tell R to which *if* the *else* belongs to.

You can also have multiple statement tests. If more than one statements can be satisfied the first that got evaluated will be executed:
```{r}
x <- 1
if(x > 0) {
  print("X is 0, or maybe not?")
} else if(x == 1) {
  print("X is 1")
} else {
  print("X is something else")
}
```

You can also chain *if-else* statements:
```{r, eval=FALSE}
x <- 5
if(x > 0) {
  if(x > 3) {
    if(x == 5) {
      cat("X is 5")
    } else {
      cat("X is greater than 3")
    }  
  } else {
    cat("X positive and less than 3")
  }
} else {
  cat("X less than 0")
}
```

R also supports switch statements.

## For loops

R supports looping through the traditional *for* and *while* statements and with the use of the *apply family* functions.

Syntax:
```
for(index in sequence) {
  do something
}
```

Again, if the To-Do step is a one-liner, curly brackets can be dropped
```{r}
a <- NULL
for(i in 1:10) a <- c(a, sum(1:i))
a
```


A sequence can be any vector, including lists
```{r}
myList <- list(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
a <- integer(10)
for(i in myList) a[i] <- sum(1:i)
a
```

## R Markdown

* This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>. 

* Simply go to File --> New File --> R Markdown

* Select PDF and you get a template.

* You most likely won't need more commands than in on the first page of [this cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf).

* When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.  

## Summary

By now, you should be able to answer the following questions:

* Why do we say that data.table is an enhanced data.frame?
* How to subset by rows or columns? Remember: DT[i, j, by].
* How to add columns?
* How to make operations with different columns?
* Which are the different types of merging?

Even if you were able to answer them, practice:

* Check all vignettes and reference manual from <https://cran.r-project.org/web/packages/data.table/>
* A really concise cheat sheet: 
* <https://s3.amazonaws.com/../assets.datacamp.com/img/blog/data+table+cheat+sheet.pdf>

## Data.table resources

<https://cran.r-project.org/web/packages/data.table/>

<https://s3.amazonaws.com/../assets.datacamp.com/img/blog/data+table+cheat+sheet.pdf>

<http://r4ds.had.co.nz/relational-data.html>

<http://adv-r.had.co.nz/Environments.html>

## Useful references

* These slides are mostly based on Hadley Wickham's book [Advanced R](http://adv-r.had.co.nz/)

* [**Everything else** from Hadley Wickham](http://hadley.nz/)

* In-depth documentations:

  * [Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf)

  * [R language definition](https://cran.r-project.org/doc/manuals/r-release/R-lang.pdf)

  * [R Internals](https://cran.r-project.org/doc/manuals/r-release/R-ints.pdf)

* Last but not least:

  * [Stackoverflow](http://stackoverflow.com/questions/tagged/r)
  


\pagebreak
