# Appendix 3: Additonal plotting tools

## Plotting themes

Themes control non-data parts of your plots:
* Complete theme
* Axis title
* Axis text
* Plot title
* Legends

They control the appearance of you plots: size, color, position... 

But nothing to do with how the data is represented.

```{r, lec07_theme_bw, fig.height=5, fig.width=6}
ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_point() +
  theme_bw()
```

Other complete themes: `theme_classic`, `theme_minimal`, `theme_light`, `theme_dark`.

More themes with `ggthemes` package (from github)
```{r, lec07_theme_ggtheme, fig.height=4.5, fig.width=5.5}
library(ggthemes)
ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_point() +
  theme_wsj() + scale_colour_wsj("colors6", "")
```

wsj: Wall Street Journal

## Axis titles
`axis.title`

`axis.title.x`

`axis.title.y`

```{r, lec07_theme_axis2, fig.height=4.5, fig.width=6}
ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_jitter() +
  theme(axis.title = element_text(size=20, color = 'red'))
```

or 
```{r, lec07_theme_axis3, fig.height=4.5, fig.width=6}
ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_jitter() +
  theme(axis.text = element_text(size=20, color = 'blue'))
```


## Summary axis elements 

(don't have to know them by heart)
The axis elements control the apperance of the axes:

Element             | Setter            | Description               
--------------------|-------------------|---------------------------
axis.line           | `element_line()`  | line parallel to axis (hidden in default themes)
axis.text           | `element_text()`  | tick labels               
axis.text.x         | `element_text()`  | x-axis tick labels        
axis.text.y         | `element_text()`  | y-axis tick labels        
axis.title          | `element_text()`  | axis titles               
axis.title.x        | `element_text()`  | x-axis title              
axis.title.y        | `element_text()`  | y-axis title              
axis.ticks          | `element_line()`  | axis tick marks           
axis.ticks.length   | `unit()`          | length of tick marks 

### Plot title themes

```{r, lec07_theme_axis4, fig.height=5, fig.width=6}
ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_jitter() +
  ggtitle('CPI vs HDI') +
  theme(plot.title = element_text(size=20, face = 'bold'))
```

Some elements affect the plot as a whole:

Element               | Setter           | Description               
----------------------|------------------|------------
plot.background       | `element_rect()` | plot background
plot.title            | `element_text()` | plot title
plot.margin           | `margin()`       | margins around plot



### Legend themes (don't have to know them by heart)
```{r, lec07_theme_axis, fig.height=5, fig.width=7}
base <- ggplot(ind, aes(CPI, HDI, color=region)) +
  geom_jitter() 
base + theme(
  legend.text = element_text(size = 15),
  legend.title = element_text(size = 15, face = "bold")
)
```

The legend elements control the apperance of all legends. You can also modify the appearance of individual legends by modifying the same elements in `guide_legend()` or `guide_colourbar()`.

Element             | Setter                    | Description                                 |
--------------------|---------------------------|---------------------------------------------|
legend.background   |  `element_rect()`         | legend background                           |
legend.key          |  `element_rect()`         | background of legend keys                   |
legend.key.size     |  `unit()`                 | legend key size                             |
legend.key.height   |  `unit()`                 | legend key height                           |
legend.key.width    |  `unit()`                 | legend key width                            |
legend.margin       |  `unit()`                 | legend margin                               |
legend.text         |  `element_text()`         | legend labels                               |
legend.text.align   |  0--1                     | legend label alignment (0 = right, 1 = left)|
legend.title        |  `element_text()`         | legend name                                 |
legend.title.align  |  0--1                     | legend name alignment (0 = right, 1 = left) |



